Brief Announcement: Tight Bounds for Online Balanced Repartitioning

### 2020-05-21 podc_NEW/podc_ba.tex
wip

We consider two flavors of the problem
in this paper. In the first one, which
was also studied by Henzinger et al.~\cite{sigmetrics19_partitioning}
(SIGMETRICS 2019) and which
we here mainly consider for the sake of a stronger
lower bound, we assume that $\sigma$
simply reveals the edges of a static graph
that can be perfectly partitioned
(i.e., in principle, no external requests
are required). 
In the second one, we consider a general
model where $\sigma$ can be arbitrary;
this was the original problem introduced
by Avin et al.~\cite{repartition-disc} at DISC 2016.
For simplicity, we will refer to the former
model as the \emph{learning} model (as
one has to learn the static communication graph) 
and to the latter as the \emph{online} model.

@@

In this section, we consider the learning variant of dynamic balanced graph partitioning problem.
For this setting, we show a~surprisingly high lower bound of $\Omega(k \cdot \ell)$ for $k\geq 3$.
The lower bound holds also in the general partitioning model (studied in Section~\ref{sec:part}).
At the end of this section, we discuss an asymptotically optimal upper bound for the learning variant.

@@

We provide a~lower bound $\Omega(k\cdot \ell)$ for the competitive ratio of any deterministic online algorithm for the learning problem.
Later, we elaborate on how to efficiently transform it to a~lower bound for the general partitioning problem.
The lower bound requires $k\geq 3$.
In contrast, for $k=2$ the learning problem is trivial: immediate collocation of communicating pairs is $1$-competitive.
In contrast,
the general partitioning problem for $k=2$ is non-trivial (see Section~\ref{sec:k2}).
%, a~lower bound of $3$ is known~\cite{repartition-disc}, and we provide a~$6$-competitive algorithm, see Section~\ref{sec:k2}.)

Throughout this paper, we often refer to groups of communicating nodes.
We use this concept slightly differently in the lower bound than the upper bounds.
In our algorithms, we group nodes with a~communication history into \emph{components}.
In this section,
we group nodes that may ever communicate, into \emph{ground sets}.

@@

The adversary constructs ground sets depending on the choices of a deterministic online algorithm.
Once we construct a ground set, it lasts until the end of the input sequence.
We start by constructing a~ground set of size $k-1$ on an arbitrarily chosen cluster.
In any partition, there must exist an isolated node collocated with the ground set of size $k-1$.
%, and we refer to it as the \emph{pivot} node.
We issue requests between this node and some node that was initially collocated with it.
By repeating such requests, almost every node is once collocated with the first ground set.
In comparison, we show that there exists an optimal offline algorithm \OPT that performs only two node exchanges ("swaps").
%Figure~\ref{fig:lb} illustrates the constructs used in the lower bound.

@@

We present an asymptotically optimal algorithm for the learning problem.
The algorithm  collocates  a~pair as soon as they communicate and it never separates them.
In order to preserve collocated pairs,
we employ the concept of components,
introduced by Avin~et~al.~\cite{repartition-disc}.

We maintain subsets of frequently communicating nodes as \emph{components}.
Initially,
each node constitutes a single-node component which we refer to as a \emph{singleton} component, 
and the node in such component is an \emph{isolated} node.
We define larger components in terms of smaller components.
Concretely,
given a (sub)sequence of requests,
two components $C_1$ and $C_2$
 \emph{merge} into one component as soon as
 for some pair of nodes $v_1 \in C_1$ and $v_2 \in C_2$,
the frequency of requests $\{v_1,v_2\}$ reaches a certain threshold through the sequence.
We keep all nodes of a~component always collocated in the same cluster,
i.e., when we  move a~node,
we move the whole component that contains it.
A~partition that has every component collocated 
is a~\emph{component respecting} partition.
%Intuitively, a component preserves the history of requests between its nodes.

In addition,
we maintain a~balanced partition of our components as long as such partition exists,
a reminiscent of partitioning given integers into sets of equal sum~\cite{integer-partitions-book}.
In contrast, our partition is time-varying:  two components are merged into one component once they communicate, and we adjust the partition accordingly.

The algorithm in this section and the algorithm for $k=3$ (cf. Section~\ref{sec:k3}) are modified versions of the algorithm DET from~\cite{repartition-disc}.
The difference is in the choice of partition after a~component merge.
In DET, the partition was arbitrary.
In the algorithm for the learning model,
we choose a~component respecting partition closest to the initial partition.
In the algorithm from Section~\ref{sec:k3}, we choose the partition closest to the current partition (a~repartition of minimum cost).
%A generic sketch of these two algorithms can be found in the appendix (Algorithm \ref{alg:ppl}).

\noindent
\textbf{Perfect Partition Learner algorithm.}
Now we describe the algorithm \PPL.
On each inter-cluster request $\{u,v\}$, 
\PPL creates new components by merging
 the two components that contain nodes $u$ and $v$.
In order to collocate nodes of the new component,
\PPL moves to a~component respecting partition that minimizes the distance to the initial partition $P_I$.
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

Fix the initial partition
$P_I :=\{ I_1, \dots, I_{\ell}\}$ and \OPT's final partition
$P_F := \{F_1, \dots, F_{\ell}\}$.
The \emph{distance} of a~partition
 $P = \{C_1, \dots, C_{\ell}\}$ from the initial partition,
defined as 
$\Delta(P) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$,
 is the number of nodes in $P$ that do not reside in their initial cluster.
In other words,
at least $\Delta(P)$ node migrations are required in order to reach the partition $P$ from $P_I$, and thus
$\OPT \geq \Delta(P_F) $.

\PPL never moves to a~partition that is more than $ \Delta^*:= \Delta(P_F) $   migrations away from~$P_I$.
This invariant latter ensures us that \PPL does not pay too much while recovering $P_F$.
We emphasize that a~\emph{repartitioning} by \PPL replaces the current partition $P$ with a~perfect partition closest to $P_I$.
This way \PPL never moves to a~partition beyond the distance $\Delta^*$.

\begin{property} \label{prop:dist<OPT}
	Let $P$ be any partition chosen by \PPL at any time.
	Then, $\Delta(P) \leq \Delta^*$.	
\end{property}

\begin{lemma}	\label{lemma:rebalancecost}
	The cost of each repartitioning by \PPL is $2\cdot\OPT$.
\end{lemma}

\begin{theorem}	\label{thm:upperbound}
	\PPL reaches the final partition $P_F$
	 and it is $(2\cdot (k-1)\cdot\ell)$-competitive.
\end{theorem}

>>>>

@@

In this section, we study a~\emph{learning} variant of dynamic balanced graph partitioning,
where the communication pattern is \emph{static}:
whether a~pair of  nodes ever communicate or not, 
is determined a~priori and is unknown to algorithms,
 and such pairs communicate forever.
Any algorithm must eventually collocate  pairs of communicating nodes,
as otherwise it cannot be competitive.
As in Henzinger et al.~\cite{sigmetrics19_partitioning}, we assume that the communication graph admits a~\emph{perfect partition},
i.e., a~partition in which no inter-cluster request ever occurs.
%any algorithm must learn this perfect partition.
The algorithm's objective is to \emph{learn} the (static) communication graph
 while serving all requests,
and without executing too many migrations.

@@

We provide a~lower bound $\Omega(k\cdot \ell)$ for the competitive ratio of any deterministic online algorithm for the learning problem.
The lower bound requires $k\geq 3$.
In contrast, for $k=2$ the learning problem is trivial: immediate collocation of communicating pairs is $1$-competitive.
%, a~lower bound of $3$ is known~\cite{repartition-disc}, and we provide a~$6$-competitive algorithm, see Section~\ref{sec:k2}.)

@@

A~\emph{ground set} is a group of nodes that communicate if they are not collocated.
The adversary constructs ground sets depending on the choices of a deterministic online algorithm.
We start by constructing a~ground set $B$ of size $k-1$ on an arbitrarily chosen cluster.
In any partition, there must exist an isolated node collocated with the ground set of size $k-1$.
%, and we refer to it as the \emph{pivot} node.
We issue requests between this node and some node that was initially collocated with it.
By repeating such requests, almost every node is once collocated with the ground set $B$.
The cost of the algorithm is linear in terms of the number of nodes, and there exists an optimal offline algorithm \OPT that performs only two node exchanges.

@@

There exists an asymptotically optimal algorithm for the learning problem.
The algorithm  collocates  a~pair as soon as they communicate and it never separates them.
To preserve collocated pairs,
we employ the concept of components,
introduced by Avin~et~al.~\cite{repartition-disc}.

\noindent
\textbf{Perfect Partition Learner algorithm.}
On each inter-cluster request $\{u,v\}$, 
\PPL  merges
 the two components that contain nodes $u$ and $v$.
To collocate nodes of the new component,
\PPL moves to partition where no component is split, that minimizes the distance to the initial partition $P_I$.

\begin{lemma}	\label{lemma:rebalancecost}
	The cost of serving each request is at most $2\cdot\OPT$.
\end{lemma}

Let $\Delta$ be the cost of \OPT.
\PPL never moves to a~partition that is more than $ \Delta$   migrations away from~the initial configuration.

\begin{theorem}	\label{thm:upperbound}
	\PPL is $(2\cdot (k-1)\cdot\ell)$-competitive.
\end{theorem}
